---
title: "Practical Machine Learning Project"
author: "Alan C Bonnici"
date: "Wednesday, November 12, 2014"
output: html_document
---

Assignment

The goal is this project is to predict the manner in which users of devices such as the *Jawbone Up*, *Nike FuelBand*, and *Fitbit* did exercise. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants who were asked to perform barbell lifts correctly and incorrectly in 5 different ways was utilised. The data and further information about the source of this data was obtained from the website: [http://groupware.les.inf.puc-rio.br/har] (see the section on the Weight Lifting Exercise Dataset). The training and testing data were downloaded from [https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv] and [https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv] respectivley.

The assignment is to create a report describing how the model was built, how cross validation was used, and the expected out of sample error. Also included are the choices that led to the conclusions.

Getting and Cleaning Data

```{r}
if (!file.exists("data")) {
    dir.create("data")
}

fileRData <- "data/predmachlearn-007.RData"
if (!file.exists(fileRData)) {
    for (fileUrl in c("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv", "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv")) {
        download.file(fileUrl, paste0("data/",basename(fileUrl)))
    }
    
    dateDownloaded <- date()
    
    # load the training data
    trainData <- read.csv("data/pml-training.csv", na.strings=c("<NA>", "NA",""))
    # load the testing data
    testData <- read.csv("data/pml-testing.csv", na.strings=c("<NA>", "NA",""))
    
    # save the data to and R object. This ensures that if the source data is deleted or altered the results are reproducable.
    save(dateDownloaded, trainData, testData, file = fileRData)
} else {
    load(file = fileRData)
    print(paste0("Using data originally downloaded on the ", dateDownloaded))
}
```

# The training data

The data consisted of a large number of fields containing NA and *null* values. These have been set to appear as NA values. The training data is made up of `r nrow(trainData)` rows and `r ncol(trainData)`.  

```{r}
# set the seed for reproducability
set.seed(123)

# load the libraries
library(caret)
```

# Cleaning the data

Cleaning the data serves the following purposes:
* It makes analysing the data more efficient in terms of processing speed;  
* It allows one to get a better understanding of the problem at hand;
* It makes it easier to figure out those columns that can be used as predictor variables and improves accuracy of the model.

The first 2 columns of the data consists of textual data that will not contribute tot he analysis. These have been removed.

```{r}
trainData <- trainData[, -(1:2)]
```


```{r}
nzv <- nearZeroVar(trainData)
trainData <- trainData[, -nzv]
```

In order to better be able to process the data those columns whose data does not change appreciably will be removed. These columns do not contribute appreciably to the model. `r length(nzv)` columns were removed from the original data.

Given that many fields consists of NAs these will be removed if they account to more than 80% of the data in that column.

```{r}
threshold <- nrow(trainData) * 0.80
colsNA <- sapply(trainData, function(x) { sum(is.na(x)) > threshold})
trainData <- trainData[, -colsNA]
```
